---
title: "Scoring the SF-36"
author: "Dom Grisafe"
date: "12/11/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#R packages
library(readxl)
library(dplyr)

#set working directory
knitr::opts_knit$set(root.dir = normalizePath("/Users/USC/Desktop/SF-36"))
#show working directory path
getwd()
```

#SF-36
The [SF-36](https://www.rand.org/health/surveys_tools/mos/36-item-short-form/survey-instrument.html), produced by RAND, is one of the most common instruments for measuring quality of life. It is frustrating to learn to manually score the SF-36, when it is clear that numerous researchers must be recoding the survey each time. This is my attempt to automate the administration and scoring of the SF-36.

##Google Form Version of SF-36
Here is the [Google Form version of the SF-36](https://goo.gl/forms/DYGom6XPEv2HlANA3). I do not know how to make a downloadable file of this form I will update the link once I lean to do so.  
Google forms eliminate data entry steps, make the data collection process more repeatable, and can save money on unnecessary administrative costs. Google forms are not commonly used in academia despite [commentaries encouraging the use of Google Forms in health and social sciences](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3884902/).

##Loading the SF-36 from Google Form to Excel to R
Once data has been collected using the Google Form, download the file as an excel document. This can then be loaded into R using the [readxl](https://www.rdocumentation.org/packages/readxl/versions/0.1.1) package. 
```{r echo = TRUE}
#import from excel to R using readxl
data <- read_excel("DataSF36.xlsx")

#look at item variables and number of observations
glimpse(data)
```

##Recoding the SF-36
RAND provides [Scoring Instructions](https://www.rand.org/health/surveys_tools/mos/36-item-short-form/scoring.html). [SAS programs](https://www.rand.org/pubs/drafts/DRU1437.html) have been published to score the SF-36 automatically, which will be used as a template for scoring the survey in R. All variables should be renamed and certain items must be reversed.
```{r recode}
#rename the columns
col <- 
#figure out how to score from Ron Klein instructions
```
#References
Figure out how to write an endnote [bibliography](https://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html) in Rmd.